Time invariant analysis of epidemics with \pkg{EpiCompare}

### 2021-05-10 jss/EpiCompare-paper/EpiCompare-paper.Rmd
pass 1 at section 3 edits

In this section, we present the tools implemented in \pkg{EpiCompare} and explain how they aid in the  data analysis pipeline.  In Fig. \ref{fig:pipeline2}, we illustrate how our package's functions fit into the data analysis pipeline introduced in Fig. \ref{fig:pipeline}.  All front-facing functions are aimed to be as user-friendly as possible. We also focus on providing the user "tidyverse" style functions, that encourage piping objects from one function to the next and follow clear "verb" naming schemes \citep{Wickham2019}. Although users can incorporate \pkg{EpiCompare} into any step in the data analysis pipeline, there are two primary points of entry.  The first point of entry is the very beginning with pre-processing and visualizing raw data, and the second point of entry is after modeling and simulation. Figure \ref{fig:pipeline2} captures these different paths, and we highlight\footnote{\ben{[Ben says: we need to make sure we actually do highlight ]}} both approaches and how to leverage \pkg{EpiCompare} in the subsections below.

@@

The first step of most data analysis is "cleaning" the data \shannon{to a format that is friendly for both computers and programmers}\footnote{\ben{[Ben says: I'm unsure this is needed and naturally adds more text. Willing to accept...]}} so it can be explored. \sout{There are multiple ways to collect epidemiological data.} \ben{In epidemology, there are multiple different formats the data can arrive in.} \footnote{\shannon{[Shannon suggests: "Epidemiological data are collected in many different formats."]} \ben{[Ben says: I see the point you're trying to make and made a new edit - but it's less connected to the previous sentence since "format" is the end.]}} Sometimes individual records are collected, with times of different states of the epidemic (infection, recovery, etc.) as well as individual information like network structure, location, and sub-population information. Other data collections focus on aggregate counts of individuals in each epidemic state.  In fact, many times only the number of new infections at each time step (e.g. weekly case counts) is observed.  Compartment totals (amounts of individuals in each state) are then imputed from those case counts\sout{ along with}\shannon{and}\footnote{\ben{[Ben says: Declined. I'm unclear why we should. Is the sentence unclear? Maybe it's the way it's a compound sentence? - If so a rewrite would be better?]}} other information about the disease and the population of interest.  In \pkg{EpiCompare}, we focus on understanding the overall impact of an outbreak at the aggregate/population level, which allows for streamlined examination of overall trends of an epidemic.

To help the practitioner examine epidemics from an aggregate/ population lens, we provide a function called `agents_to_aggregate()`. This function transforms data about individual/agents' initial entry into each state (e.g. start of infection, start of recovery, etc.) to an aggregate view of how many individuals were in a state at a given time. There are often situations where \ben{\sout{grouping} aggregating}\footnote{\ben{[Ben says: reason - connect to languare from sentence before.]}} agents into subpopulations (e.g. sub\sout{populations}\shannon{ groups}\footnote{\ben{[Ben says: declined. I'm unclear why we'd do this.]}} defined by age or sex) can highlight different aggregate level trends. For example, research by \citet{rvachev1985,anderson1992,worby2015} develop state-based models that account for differing disease dynamics in different subpopulations. In \pkg{EpiCompare}, we facilitate subpopulation analysis by combining the function \pkg{dplyr}::`group_by()` and `agent_to_aggregate()` to provide aggregation \sout{at a}\shannon{by}\footnote{\ben{[Ben says: declined. unclear about this change. Also would it not be "by groups"? Orignial seems to flow more.]}} group level.

The `agents_to_aggregate()` function is flexible and can deal with a wide range of information about each individual. 
\ben{\sout{It can,} In fact, this function can}\footnote{\shannon{[Shannon suggests: "Theoretically, this agents_to_aggregate() can"]} \ben{[Ben says: I just removed the "theoretically" since it seems to suggest more than we want. Also I didn't see your original edit. both edits connect to the previous sentence better.]}} account for infinitely many states.  This functionality allows the practitioner to aggregate information relative to common states (e.g. "Susceptible", "Infectious", and "Recovered") as well as more complex states (e.g. "Exposed", "iMmune", "Hospitalized"). Additionally, `agents_to_aggregate()` permits indicators for death/exit and birth/entry dates. Overall, this function is a powerful tool for pre-processing data, \shannon{and it lowers the barrier for entry into data analysis for less experienced practicioners.}\footnote{\shannon{[Shannon says: I added this in theme with 'highlighting a point of entry']} \ben{[Ben says: I'm unclear about this comment but I'm find with the addition...]}}

@@

\ben{\sout{With raw data, "Getting to know" our} In the early stages of a project, getting to know the}\footnote{\shannon{[Shannon suggested: "Getting to know" our]} \ben{[Ben says: even though I was the original one with the quotes - one should always stray away from it. I kept the implicit connection to the section we're in given I'm not sure people are reading the titles super well.]}} data currently means figuring out useful combinations of visualizations \ben{\sout{,}and} numerical summaries and \sout{\sout{subsets}\shannon{groupings}}\ben{exploring different groupings of the data}\footnote{\ben{[Ben says: the action "figuring out" and "groupings" was a bit unclear what actually was going to happen - so I changed it.]}}. \ben{\sout{An expert coder has many ways to successfully explore the data in an aggregate lens using }}`agents_to_aggregate()`\ben{\sout{. For less experienced coders, \pkg{EpiCompare} also includes tools to rapidly explore data that has three epidemic states.} An expert coder can start from } `agents_to_aggregate()` \ben{to successfully accomplish EDA in many ways, but \pkg{EpiCompare} also includes tools that allow a novice coder to rapidly explore data, as long as there three unique epidemiological states (like the SIR model).}\footnote{\ben{[Ben says: I'm reverting to the old text - it's much clearer. - happy to have a a discussion on it. The crossed out text didn't capture what an expert coder was really going to do and how that differed.]}} Building on \ben{\sout{the tools in}} \pkg{ggplot2} and \pkg{ggtern} \ben{packages}, \pkg{EpiCompare}'s `geom_aggregate()` provides a rapid way to explore different subpopulations' experiences of an epidemic \citep{Wickham2016, Hamilton2018}. \ben{The function }`geom_aggregate()` \ben{provides a visualization tool to holistically examine aggregate level information across different subpopulations by visualizing each subpopulation's epidemic trajectory in the three-dimensional state space.}  \footnote{\ben{[Original text: "By combiing the ideas behind agent2aggregate for three-state models to examine subpopulation trajcetories in 2d simplex space."]} \shannon{[Shannons says: I got rid of the end of this sentence because I think it was putting multiple ideas in one sentence.]}\ben{[Ben says: I'm unclear why this warrants just deleting it. I've provided a newly written sentence...]}}  Visualization tools for three-state models were developed because (1) SIR models are some of the most common and basic epidemic state-based models and (2) our three-dimensional simplex representation of these epidemics emphasizes a "time-invarance" representation of the data (for a refresher see Section \ref{sec:time-invariant}).

@@

\ben{[Ben says: think about this section and if it highlights that we can bring in outside models...]}

After getting a good sense of what a past or current epidemic looks like with EDA, the next step is often model fitting and/or simulations. \sout{In this step and the next step (post-processing), we discuss how to \shannon{easily include external models and simulations originating outside into the \pkg{EpiCompare} data analysis pipeline.}}\ben{In this step and the next step (post-proessing), we highlight how practitioners can pair models and simulations of epidemics from outside of \pkg{EpiCompare} with analysis and simulation tools in \pkg{EpiCompare}.}\footnote{\ben{[Ben says: the original rewrite didn't seem to clear - specifically in how the practitioner would be using outside models in this step.]}} While \pkg{EpiCompare} does not focus on fitting model(s) to data, we do provide some flexible functions for simulation of basic discrete-time epidemic-state models. \ben{These functions simulation individual-level information based on practitioner estimated transition rates between states and} can be combined with `agents_to_aggregate()` to view these simulations through an aggregate lens.  The function `simulate_SIR_agents()` simulates a basic SIR epidemic with user inputs for the number of simulations, the initial number in each state, the infection and recovery parameters $(\beta, \gamma$), and the total number of discrete time steps.  This function allows for easy access to SIR model analysis and comparison.  Beyond SIR models, the function `simulate_agents()` takes as input a user-specified state-transition matrix and other epidemic parameters to allow the user to create simulations for an outbreak with \textit{any} number of states and any number of transitions among them.  This flexibility in states can be used to also reflect group-based dynamics.  In turn, this allows for users to explore the space of models in an intuitive way without getting bogged down by too much mathematical detail. For consistency, we have made output from `simulate_agents()` and `simulate_SIR_agents()` compatible with `agents_to_aggregate()` so aggregate information may easily be accessed.

@@

\ben{[Ben says: I think we should remind the reader that we care more about simulations, in order to compare fitted models between themselves and the true epidemics.]}

\ben{[Ben says: this replaces the first paragraph below]}
If the practitioner wishes to compare models-to-observations or even models-to-models, they needs to post-process their models and simulations to disseminate the results in an easily digestable format. In general, post-processing of modeling and simulation consists of making summary statistics, plots, and tables. \ben{Model \sout{The}} summaries can be very complex, and as a result, a number of epidemic modeling \\proglang{R} packages return \ben{\sout{a}} special class \ben{objects}.  The special class\ben{\sout{es} objects} often contain a plethora of information from residuals, model diagnostics, input parameters, and more.  While incredibly useful, these special classes can be difficult for novice coders to \ben{\sout{navigate} handle}.

<!-- [\ben{[Ben says: replaced with above paragraph]} Post-processing of modeling and simulation consists of making summary statistics, plots, tables, and other ways to disseminate information to the public. For example, comma separated value files (`.csv`) are a  standard way to share information within tables.  However, model output is often far more complicated than what a traditional `.csv` would allow.  \shannon{Do we need three sentences about csvs?} As a result, a number of epidemic modeling packages return a special class, specific to their modeling.  The special classes often contain a plethora of information from residuals, model diagnostics, input parameters, and more.  While incredibly useful, these special classes can be difficult for novice coders to navigate. ] -->

@@

\shannon{[Shannon says: Here Ben talks about filament compression and tidy\_dist mats, etc.]} \ben{[Ben says: not currently thinking about including tidy\_dist stuff - will need to think about this as it is including in the pipeline 2 image...]} \footnote{\ben{[Ben says: are these comments still relevant?]}}

To utilize simulations of epidemics in later time-invariant analysis we also provide a function to convert temporally defined epidemics to filamental representations. Specifically, we provide the function `filament_compression()` to convert simulation(s) to filaments as expressed by presenting the epidemic as a ordering of some common fixed number of points so that they are equally spaced along the original path in the proportional state space.

\shannon{[Shannon says: Shannon's attempt at above paragraph is below..  I tried to more smoothly connect to previous paragraph and then kinda copped out at trying to reword the filament description by just referring back to the previous section.]}

\pkg{EpiCompare} also provides a tool to convert time-dependent epidemic simulations into their time-invariant filamental representations.  Specifically, `filament_compression()` converts simulations to filaments (see Section \ref{sec:time-invariant}) \ben{\sout{so practitioners can view model simulations and results through a time-invariant lens.}which allows practitioner to later apply time-invariant tools to these compressed epidemiological objects.}\footnote{\ben{[Ben says: I worry about this statement - it is very broad / seems to claim a lot but not very clearly...]}}  These tools were developed to provide another natural entry point into the \pkg{EpiCompare} data analysis pipeline for situations where modeling and analysis is already completed and practitioners are looking for simple and transparent tools to help understand and disseminate results.

@@

\ben{[Ben says: I didn't look at this section - it looks like your comments are more "let's" talk about this and not suggestions / feedback.]}

<!-- \ben{[Brainstorming: (intro) discuss model vs model comparison tools, (a) highlight that standard comparisons can be done if models share metrics or parameters (and similarly some comparisons can be done to the real epidemics). (b) Present time-invariance tools that allow us to compare more types of models (and can also give new tools to assess models beyond the standard ones). (1) introduce prediction regions (a) geom_prediction_bands,  (2) discuss different extremeness measures.]} -->

\shannon{Finally, \pkg{EpiCompare} can be used the last step of the data analysis pipeline with its comparison and assessment utility functions.}  As introduced in Section \ref{sec:beyond-r0-sir} there's a \sout{lot of}\shannon{much} potential for time-invariant tools to help compare and assess epidemics and models/simulations. In \pkg{EpiCompare} we provide a set of comparison and assessment tools for models that extend beyond the standard performance metrics (e.g. mean squared error or AIC) and focus on assessing the structural information the models capture. This approaches work well on models where \sout{online} one "cycle" of the epidemic has occurred (no recovered individuals have been susceptible again)\footnote{\ben{[Ben says: Shannon - do you think this is clear / a desirable way to define this - we define it slightly differently in section 2.2.]}\shannon{Shannon says I don't think I understand what you are trying to say there.  We should chat about it.}}. Epidemics are complex objects, and we provide tools to create prediction regions \shannon{with differing desired characteristics}\footnote{\shannon{Shannon says: I want to connect epidemics being complex to the fact that prediction is hard and not all prediction regions tell you the same thing}} from simulated epidemics. For three-state epidemic models, we provide the `ggplot`/`ggtern` extension `geom_prediction_band()` which creates a prediction region around the top $1-\alpha$ proportion [\shannon{good place for Ben paper cite?}] of the simulations (where the simulations treated as filaments). In \pkg{EpiCompare} we also provide these prediction regions for epidemic models with with more than three states. The functions `create_convex_hull_structure()` and `create_delta_ball_structure()` create different geometric representations of prediction regions for any state-based model. For both of these geometric objects, we provide functions to check if a path is contained (`contained()`) and the ability to assess the Hausdorff distance between prediction regions based on simulations from different model (`hassdorf_dist()`).  \footnote{\ben{[Ben says: should we discuss the details of geom\_prediction\_bands more? ]}\shannon{[Shannon says: I don't think so, best left to the vignette and data example]}}

We also provide functions to calculate the "extremeness" of a true epidemic [\shannon{trajectory?}] compared to simulated epidemics via the equi-distance filamental trajectory representation as mentioned in Section \ref{sec:beyond-r0-sir}. Specifically, functions like  `distance_pseudo_density_function()` can calculate a pseudo-density estimate of the true epidemic relative to simulated ones. Functions `distance_depth_function()` and `local_distance_depth_function()` provide depth scores that suggest how geometrically central an epidemic is to simulations. \footnote{\ben{[Ben says: should we define these more in this paragraph? Introducted in 2.2. as well and not defined.]}\shannon{[Shannon says: I'd throw a citation or two but again limit the detail.]}}

\ben{[Ben says: Replaced with paragraphs above.]} Comparison and assessment of model fit or comparisons of one model to another model can be performed in a variety of ways including mean square error, AIC, plots, and more.  Perhaps the most useful tool \pkg{EpiCompare} has to offer to the expert, for comparison and assessment of models, is in its post-processing tools which create a standard output.  It is then a matter of writing a script or function made for that standard output to assess the results from multiple models in the way the user desires\footnote{\ben{[Ben says: Shannon, doesn't this below in the previous section?, I might drop this paragraph except for the introduction sentence - is it really true that the standardization is most valuable?]}\shannon{I agree with you now that this section is nice and filled in.  Let's drop this paragraph.}}.

\ben{[Ben says: Replaced with paragraphs above.]} However, for those who like more concrete tools, \pkg{EpiCompare} offers functions to compare prediction regions to one another including `geom_prediction_band()` (which plots the region), and `create_{convex_hull,delta_ball}_structure()`(which returns the `R` output for the given structure), and `contained()` (which allows the user to determine if one set of points is contained in a prediction band).  Additionally, we offer ways to determine if model outputs are compatible with one another, that is how extreme one output is to another.  \shannon{Ben says something about distance}

>>>>

In this section, we present the tools implemented in \pkg{EpiCompare} and explain how they aid in the  data analysis pipeline.  In Fig. \ref{fig:pipeline2}, we illustrate how our package's functions fit into the data analysis pipeline introduced in Fig. \ref{fig:pipeline}.  All front-facing functions in \pkg{EpiCompare} are aimed to be as user-friendly as possible. We also focus on providing the user "tidyverse" style functions, that encourage piping objects from one function to the next and follow clear "verb" naming schemes \citep{Wickham2019}. Although users can incorporate \pkg{EpiCompare} into any step in the data analysis pipeline, there are two primary points of entry.  The first point of entry is the very beginning with pre-processing and visualizing raw data, and the second point of entry is after modeling and simulation. Figure \ref{fig:pipeline2} captures these different paths, and we highlight how to leverage \pkg{EpiCompare} functionalities in the subsections below.

@@

The first step of most data analysis is "cleaning" the raw data so it can be explored. Before data can be explored, they must be collected.  Raw data can appear in many different formats. Sometimes individual records are collected, with times of different states of the epidemic (infection, recovery, etc.) as well as individual information like network structure, location, and sub-population information. Other data collections focus on aggregate counts of individuals in each epidemic state.  In fact, many times only the number of new infections at each time step (e.g. weekly case counts) is observed.  Compartment totals (amounts of individuals in each state) are then imputed from those case counts other information about the disease and the population of interest.  In \pkg{EpiCompare}, we focus on understanding the overall impact of an outbreak at the aggregate/population level, which allows for streamlined examination of overall trends of an epidemic.

To help the practitioner examine epidemics from an aggregate/population lens, we provide a function called `agents_to_aggregate()`. This function transforms data about individual/agents' initial entry into each state (e.g. start of infection, start of recovery, etc.) to an aggregate view of how many individuals were in a state at a given time. There are often situations where aggregating agents into subpopulations (e.g. subpopulations defined by age or sex) can highlight different aggregate level trends. For example, research by \citet{rvachev1985,anderson1992,worby2015} develop state-based models that account for differing disease dynamics in different subpopulations. In \pkg{EpiCompare}, we facilitate subpopulation analysis by combining the function \pkg{dplyr}::`group_by()` and `agent_to_aggregate()` to provide aggregation at a subpopulation level.

\shannon{Besides aiding subpopulation analysis,} `agents_to_aggregate()` can deal with a wide range of information about each individual. 
In fact, this function account for infinitely many states.  This functionality allows the practitioner to aggregate information relative to common states (e.g. "Susceptible", "Infectious", and "Recovered") as well as more complex states (e.g. "Exposed", "iMmune", "Hospitalized"). Additionally, `agents_to_aggregate()` permits indicators for death/exit and birth/entry dates. Overall, this function is a powerful tool for pre-processing data.

@@

 In the early stages of a project, familiarizing oneself with the data usually means figuring out useful combinations of visualizations and numerical summaries of the data \shannon{both at population and subpopulation level.}  An expert coder can start with `agents_to_aggregate()` to successfully accomplish EDA in many ways. \pkg{EpiCompare} also includes tools that allow a novice coder to rapidly explore data, provided there three unique epidemiological states (like in the SIR model). Building on \pkg{ggplot2} and \pkg{ggtern} packages, \pkg{EpiCompare}'s `geom_aggregate()` provides a rapid way to explore how different subpopulations experience of an epidemic \citep{Wickham2016, Hamilton2018}. The function `geom_aggregate()` provides a visualization tool to holistically examine aggregate level information across different subpopulations by visualizing each subpopulation's epidemic trajectory in the three-dimensional state space.  Visualization tools for three-state models were developed because (1) SIR models are some of the most common and basic epidemic state-based models and (2) our three-dimensional simplex representation of these epidemics emphasizes a "time-invarance" representation of the data (for a refresher see Section \ref{sec:time-invariant}).

@@

After getting a sense of what a past or current epidemic looks like with EDA, often the next step in the data analysis pipeline is model fitting and/or simulation. In this step and the next step (post-proessing), we highlight how practitioners can pair models and simulations of epidemics from outside of \pkg{EpiCompare} with analysis and simulation tools in \pkg{EpiCompare}. While \pkg{EpiCompare} does not focus on fitting models to data, we do provide some flexible functions for simulation of basic discrete-time epidemic-state models. These functions simulate individual-level information based on practitioner estimated transition rates between states and can be combined with `agents_to_aggregate()` to view these simulations through an aggregate lens.  The function `simulate_SIR_agents()` simulates a basic SIR epidemic with user inputs for the number of simulations, the initial number in each state, the infection and recovery parameters $(\beta, \gamma$), and the total number of discrete time steps.  This function allows for easy access to SIR model analysis and comparison.  Beyond SIR models, the function `simulate_agents()` takes as input a user-specified state-transition matrix and other epidemic parameters to allow the user to create simulations for an outbreak with \textit{any} number of states and any number of transitions among them.  This flexibility in states can be used to also reflect group-based dynamics.  In turn, this allows for users to explore the space of models in an intuitive way without getting bogged down by too much mathematical detail. For consistency, we have made output from `simulate_agents()` and `simulate_SIR_agents()` compatible with `agents_to_aggregate()` so aggregate information may easily be accessed.

@@

If practitioners wish to compare models-to-observations or even models-to-models, they needs to post-process their models and simulations to disseminate the results in an easily digestible format. In general, post-processing of modeling and simulation consists of making summary statistics, plots, and tables. Model output summaries can be very complex, and as a result, a number of epidemic modeling \proglang{R} packages return a special class.  The special classes often contain a plethora of information about residuals, model diagnostics, input parameters, and more.  While incredibly useful, these special classes can be difficult for novice coders to handle.

@@

To utilize simulations of epidemics in  time-invariant analysis, we also provide a function to convert temporally defined epidemics to filamental representations. Specifically, we provide the function `filament_compression()` to convert simulation(s) to filaments as expressed by presenting the epidemic as an ordering of some common fixed number of points so that they are equally spaced along the original path in the proportional state space. These tools were developed to provide another natural entry point into the \pkg{EpiCompare} data analysis pipeline for situations where modeling and analysis are already completed and practitioners are looking for simple and transparent tools to help understand and disseminate results.

@@

Finally, in \pkg{EpiCompare} we provide a set of comparison and assessment tools for model and simulation results that extend beyond the standard performance metrics (e.g. mean squared error or AIC). Corresponding to Section \ref{sec:beyond-r0-sir}, \pkg{EpiCompare} provides a set of time-invariant tools to compare and evaluate epidemic models and simulations. We have found these tools to be specifically applicable for situations where only one "season" or "cycle" of the epidemic has occurred.

One tool we provide to assess models is through the creation of geometric prediction regions, which are useful when we treat epidemics like filaments. If there is a set of simulated epidemics from a model, we can create a geometric prediction region for the expected trajectory of the epidemic in the state space. For three-state epidemic models, we provide the `ggplot`/`ggtern` extension `geom_prediction_band()` which creates a prediction region around the top $1-\alpha$ proportion of the simulations. In this visual setting, comparing this prediction to the true epidemic trajectory or comparing the prediction regions defined by two different models' simulations can be done with the 'eye test.' In \pkg{EpiCompare} we also provide these prediction regions for epidemic models with with more than three states. The functions `create_convex_hull_structure()` and `create_delta_ball_structure()` create different geometric representations of prediction regions for any dimensional state-based model. For both of these geometric structures, we provide functions to check if a path is contained (`contained()`) and the ability to assess the Hausdorff distance between prediction regions based on simulations from different model (`hausdorff_dist()`).

We also provide functions to calculate the "extremeness" of a true epidemic trajectory compared to simulated epidemics via the equi-distance filamental trajectory representation as mentioned in Section \ref{sec:beyond-r0-sir}. Specifically, functions like  `distance_pseudo_density_function()` can calculate a pseudo-density estimate of the true epidemic relative to simulated ones. Functions `distance_depth_function()` and `local_distance_depth_function()` provide depth scores that suggest how geometrically central an epidemic is to simulations.
